---
title: Plant Phenology Data Visualization
author: "Zoe Zhou"
date: today
format: 
  html:
    code-fold: true
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
theme: minty
---
## Set up
```{r}
library(tidyverse)
library(DBI)
library(duckdb)
library(dbplyr)
```

## Establishing a connection to the database and check for tables
```{r}
conn <- DBI::dbConnect(duckdb::duckdb(), dbdir="database/phenology.db")
DBI::dbListTables(conn)
```

## Load the data
```{r}
plant <- tbl(conn, "plant")
census <- tbl(conn, "census")
```
## Research question
Is there a relationship between plant height and stem diameter across different growth forms?

## Join tables and prepare for visualization
- inner join annual census data with plant data
- filter out missing values for stem diameter and height
- group by growth form
```{r}

# Create the joined dataset 
plant_census <- inner_join(plant, census, by="individualID") %>% 
  filter(!is.na(stemDiameter), !is.na(height)) %>% 
  group_by(growthForm) %>% 
  show_query() 
```

## Plotting

```{r}
# plot
plot <- ggplot(plant_census, aes(x=stemDiameter, y=height, color=growthForm)) +
  geom_point(alpha=0.7) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Plant Height vs. Stem Diameter",
    x = "Stem Diameter (cm)",
    y = "Height (cm)",
    color = "Growth Form"
  ) +
  theme_minimal()
plot
```

